|

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.Rmd
revision to appendix

@@

#alpha for ribbon in the smooth plots
al<-0.8

thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))

@@

The first function is `example.R`, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript. Both rm-ANOVA and LMEM with interaction are fitted to the data. The error for each simulated trend can be correlated or uncorrelated as well.

@@

  #Create mean response matrix: linear or quadratic
  mu <- matrix(0, length(x), 2)
  # linear response
  if (fun_type == "linear") {
    mu[, 1] <- - (0.25*x)+2  
    mu[, 2] <- 0.25*x+2
  } else {
    # quadratic response (non-linear)

    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
  }

@@

  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
  #between-group variability in the response.
  errors <- array(0, dim = c(length(x), 2, 10))
  #create an array where 10 observations per each time point for each group are to be stored

@@

#LMEM: time and treatment interaction model, compound symmetry 
  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject)
  )

  #create a prediction frame where the model can be used for plotting purposes
  pred_dat <- expand.grid(
    treatment = factor(1:2), 
    time = unique(dat$time)
  )

@@

  #return everything in a list
  return(list(
    dat = dat,
    pred_dat = pred_dat,
    fit_anova=fit_anova,
    fit_lme = fit_lme
  ))
}
```

@@

Function `plot_example.R` uses the output of `example.R` to show the fit of a rm-ANOVA and a LMEM. This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case. The corresponding rm-ANOVA and LMEM fits are also presented, and we show the complete output in the next subsection.

@@

## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
plot_example <- function(sim_dat) {
    txt<-20
    p1 <- sim_dat$dat %>%
        ggplot(aes(x = time,
                   y = y,
                   group = treatment,
                   color = treatment)
        ) +
        geom_point(show.legend=FALSE) +
        labs(y='response')+
        geom_line(aes(x = time,
                      y = mu,
                      color = treatment),
                  show.legend=FALSE) +
        theme_classic() +
        theme(plot.title = element_text(size = txt,
                                        face = "bold"),
              text=element_text(size=txt))+
        thm1

    #plot the simulated data with trajectories per each subject
    p2 <- sim_dat$dat %>%
        ggplot(aes(x = time,
                   y = y,
                   group = subject,
                   color = treatment)
        ) +
        geom_line(aes(size = "Subjects"),
                  show.legend = FALSE) +
        # facet_wrap(~ treatment) +
        geom_line(aes(x = time,
                      y = mu,
                      color = treatment,
                      size = "Simulated Truth"),
                  lty = 1,show.legend = FALSE) +
        labs(y='response')+
        scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
        theme_classic()+
        theme(plot.title = element_text(size = txt,
                                        face = "bold"),
              text=element_text(size=txt))+
        thm1

    #plot the errors
    p3 <- sim_dat$dat %>%
        ggplot(aes(x = time,
                   y = errors,
                   group = subject,
                   color = treatment)) +
        geom_line(show.legend=FALSE) +
        labs(y='errors')+
        theme_classic()+
        theme(plot.title = element_text(size = txt,
                                        face = "bold"),
              text=element_text(size=txt))+
        thm1

    #plot the model predictions for rm-ANOVA
    p4 <- ggplot(sim_dat$dat,
                 aes(x = time,
                     y = y,
                     color = treatment)) +
        geom_point(show.legend=FALSE)+
        labs(y='response')+
        geom_line(aes(y = predict(sim_dat$fit_anova),
                      group = subject, size = "Subjects"),show.legend = FALSE) +
        geom_line(data = sim_dat$pred_dat,
                  aes(y = predict(sim_dat$fit_anova,
                                  level = 0,
                                  newdata = sim_dat$pred_dat),
                      size = "Population"),
                  show.legend=FALSE) +
        guides(color = guide_legend(override.aes = list(size = 2)))+
        scale_size_manual(name = "Predictions",
                          values=c("Subjects" = 0.5, "Population" = 3)) +
        theme_classic() +
        theme(plot.title = element_text(size = txt,
                                        face = "bold"),
              text=element_text(size=txt))+
        thm1

    #plot the LMEM predictions
    p5 <- ggplot(sim_dat$dat,
                 aes(x = time,
                     y = y,
                     color = treatment)) +
        geom_point()+
        labs(y='response')+
        geom_line(aes(y = predict(sim_dat$fit_lme),
                      group = subject, size = "Subjects")) +
        geom_line(data = sim_dat$pred_dat,
                  aes(y = predict(sim_dat$fit_lme,
                                  level = 0,
                                  newdata = sim_dat$pred_dat),
                      size = "Population")) +
        guides(color = guide_legend(override.aes = list(size = 2)))+
        scale_size_manual(name = "Predictions",
                          values=c("Subjects" = 0.5, "Population" = 3)) +
        theme_classic() +
        theme(plot.title = element_text(size = txt,
                                        face = "bold"),
              text=element_text(size=txt))+
        thm1

    if(option=='simple'){
        return((p1+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
    }
    else {
    return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
    }

@@

In this subsection, we use `example.R` and `plot_example.R` to create an expanded version of Figure 1 on the main manuscript. The difference here is that in the main manuscript `plot_example.R` uses `option='simple'` to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit. Here, we use the option `composite` to generate the additional panels for the uncorrelated errors and their respective rm-ANOVA and LMEM fits.

@@

The chunk below calls both `example.R` and `plot_example_Appendix.R` to simulate data and create the composite plots.

@@

source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example.R"))

A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),option='composite')

B1<-plot_example(example(fun_type = "linear", error_type = "independent"),option='composite')

C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"),option='composite')

D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"),option='composite')

@@

```{r, linear-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
# linear response, correlated errors (subject effect)
A1/B1+plot_annotation(tag_levels = 'A')
```

@@

```{r, quadratic-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
# linear response, correlated errors (subject effect)
C1/D1+plot_annotation(tag_levels = 'A')
```

@@

From the model, we extract the original and weighed four basis functions (and the intercept), and add these last weighed basis functions to construct the smoother.

@@

#basis functions: this script creates Fig 2. by calculating a GAM for the Group 1 data of Figure 1,
# extracting the basis functions
#and creates objects p11,p12,p13,p14 for plotting, which are combined in b_plot to create the
#final composite figure
n_time = 6
x <- seq(1,6, length.out = n_time)
mu <- matrix(0, length(x), 2)
mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
y <- array(0, dim = c(length(x), 2, 10))
errors <- array(0, dim = c(length(x), 2, 10))
for (i in 1:2) {     # number of treatments
    for (j in 1:10) {  # number of subjects
        # compound symmetry errors
        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
        y[, i, j] <- mu[, i] + errors[, i, j]
    }
}

@@

#extract  "Group 1" to fit the GAM
dat<-subset(dat,treatment==1)
#keep just the response and timepoint columns
dat<-dat[,c('y','time')]

#GAM model of time, 5 basis functions
gm<-gam(y~s(time,k=5),data=dat, method= "REML")

#model_matrix (also known as) 'design matrix'
#will contain the smooths used to create  model 'gm'
model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))

time<-c(1:6)

basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
#basis<-model_matrix[1:6,-1] #extracting basis
colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
basis<-basis %>% #pivoting to long format
    pivot_longer(
        cols=starts_with("s")
    )%>%
    arrange(name) #ordering

#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
ln<-6*(length(coef(gm)))

basis_plot<-data.frame(Basis=integer(ln),
                       value_orig=double(ln),
                       time=integer(ln),
                       cof=double(ln)
)

basis_plot$time<-rep(time) #pasting timepoints
basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
basis_plot$value_orig<-basis$value #pasting basis values
basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
basis_plot<-basis_plot%>%
    mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be
#multiplied by the coefficients

@@

basis_names<-c(
    `1`="Intercept",
    `2`="1",
    `3`="2",
    `4`="3",
    `5`="4"
)

@@

smooth<-basis_plot%>%
    group_by(time)%>%
    summarize(smooth=sum(mod_val))

#original basis
sz<-1
p11<-ggplot(basis_plot,
            aes(x=time,
                y=value_orig,
                colour=as.factor(Basis)
            )
)+
    geom_line(size=sz,
              show.legend=FALSE)+
    geom_point(size=sz+1,
               show.legend = FALSE)+
    labs(y='Basis functions')+
    facet_wrap(~Basis,
               labeller = as_labeller(basis_names)
    )+
    theme_classic()+
    thm1

#penalized basis
p12<-ggplot(basis_plot,
            aes(x=time,
                y=mod_val,
                colour=as.factor(Basis)
            )
)+
    geom_line(show.legend = FALSE,
              size=sz)+
    geom_point(show.legend = FALSE,
               size=sz+1)+
    labs(y='Penalized \n basis functions')+
    scale_y_continuous(breaks=seq(-1,1,1))+
    facet_wrap(~Basis,
               labeller=as_labeller(basis_names)
    )+
    theme_classic()+
    thm1

#heatmap of the  coefficients
x_labels<-c("Intercept","1","2","3","4")
p13<-ggplot(basis_plot,
            aes(x=Basis,
                y=Basis))+
    geom_tile(aes(fill = cof),
              colour = "black") +
    scale_fill_gradient(low = "white",
                        high = "#B50A2AFF")+
    labs(x='Basis',
         y='Basis')+
    scale_x_discrete(labels=x_labels)+
    geom_text(aes(label=round(cof,2)),
              size=7,
              show.legend = FALSE)+
    theme_classic()+
    theme(legend.title = element_blank())

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
    geom_point(size=sz+1,alpha=0.5)+
    thm1+
    labs(y='Simulated \n response')+
    geom_line(data=smooth,
              aes(x=time,
                  y=smooth),
              color="#6C581DFF",
              size=sz+1)+
    theme_classic()

#Combining all
b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
    theme(
        text=element_text(size=18)
    )

@@

The next chunk presents the code of the function `simulate_data.R`, which we use in the main manuscript to simulate data that follows the reported trends in $\mbox{StO}_2$ in the literature. Although the default standard deviation (SD) to generate the normally-distributed data in the function is 10%, we use a standard deviation of 10% in the main manuscript. We use the default number of 10 observations at each time point in the simulated data. Note that the input for the function is `dat`, needs to be a dataframe that contains the original values for the trends in each group over time from the literature. In our manuscript, we use the dataframe

@@

#data from Vishwanath et. al.
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )
```

@@

simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(
            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
            subject=rep(1:10),
            subject=factor(paste(subject, Group, sep = "-"))
        ) %>%
        ungroup()

@@

We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function `confint` from base R. The boundaries for the pairwise and simultaneous CIs are obtained by using the type `"confindence"` or `"simultaneous"` in `confint` and are stored in `ci` and `si` respectively.

@@

```{r,pointwise-plot, eval=FALSE}
#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
#appraise the model
appraise(mod_m1)

@@

f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
geom_line(lwd = 1) +
geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
inherit.aes = FALSE, alpha = 0.7,
show.legend=FALSE) +
geom_ribbon(data = si,
mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
inherit.aes = FALSE, alpha = 0.4,
show.legend=TRUE)+
    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
    geom_point(data=dat_missing, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group), 
                            size=1.5,
                        alpha=0.6, 
               inherit.aes = FALSE)+
    labs(y=expression(atop(StO[2],'(incomplete observations)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22))+
        thm1

@@

This subsection presents the code used in functions `pointwise_comparisons.R` and `difference_smooths.R`, which generate the pointwise and simultaenous CI for the estimated difference between the smooths. In essence, `pointwise_comparisons.R` computes the difference between the smooths and the pointwise CI. Then, `difference_comparisons.R` uses the output of `pointwise_comparisons.R` to estimate the simultaneous CI.

@@

Then, `difference_smooths.R` takes the output from the previous function and uses it to compute the simultaneous CI.
We present the code below.

@@

    # smooths in model
    S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
    # select smooths
    select <-
        gratia:::check_user_select_smooths(smooths = S, select = smooth,
                                           partial_match = partial_match)#,
    # model_name = expr_label(substitute(object)))
    sm_ids <- which(select)
    smooths <- gratia::get_smooths_by_id(model, sm_ids)
    sm_data <- map(sm_ids, gratia:::smooth_data,
                   model = model, n = n, include_all = TRUE)
    sm_data <- bind_rows(sm_data)
    by_var <- by_variable(smooths[[1L]])
    smooth_var <- gratia:::smooth_variable(smooths[[1L]])
    pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
                                     stringsAsFactor = FALSE))
    names(pairs) <- paste0("f", 1:2)

@@

pairwise_limits<-function(dataframe){
    #extract values where the lower limit of the ribbon is greater than zero
    #this is the region where the control group effect is greater

    v1<-dataframe%>%
        filter(lower_s>0)%>%
        select(Day)
    #get day  initial value
    init1=v1$Day[[1]]
    #get day final value
    final1=v1$Day[[nrow(v1)]]
    #extract values where the value of the upper limit of the ribbon is lower than zero
    #this corresponds to the region where the treatment group effect is greater
    v2<-dataframe%>%
        filter(upper_s<0)%>%
        select(Day)

    init2=v2$Day[[1]]
    final2=v2$Day[[nrow(v2)]]
    #store values
    my_list<-list(init1=init1,
                  final1=final1,
                  init2=init2,
                  final2=final2)
    return(my_list)

>>>>

```{r, echo = FALSE}
knitr::opts_chunk$set(tidy = TRUE)
```

@@

#alpha for ribbon in the smooth plots
al<-     0.8

thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
                           direction = -1, aesthetics = c("colour", "fill"))

@@

The first function is `example()` which is in the file named `example.R` in the `scripts/` folder, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript. Both rm-ANOVA and LMEM with interaction are fitted to the data. The error for each simulated trend can be correlated or uncorrelated as well.

@@

  #Create mean response matrix: linear or quadratic
  mu <- matrix(0, length(x), 2)

  if (fun_type == "linear") {
    # linear response
    mu[, 1] <- -(0.25*x)+2  
    mu[, 2] <- 0.25*x+2
  } else {
    # quadratic response (non-linear)
    mu[, 1] <- -(0.25 * x^2) + 1.5 * x - 1.25
    mu[, 2] <- (0.25 * x^2) - 1.5 * x + 1.25
  }

@@

  #Create array to store the "errors" for each group at each timepoint. The "errors" are the between-group variability in the response.
  errors <- array(0, dim = c(length(x), 2, 10))
  #create an array where 10 observations per each time point for each group are to be stored

@@

  #LMEM: time and treatment interaction model, compound symmetry 
  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject))

  #create a prediction frame where the model can be used for plotting purposes
  pred_dat <- expand.grid(treatment = factor(1:2), 
                          time = unique(dat$time))

@@

  #return everything in a list
  return(list(
    dat       = dat,
    pred_dat  = pred_dat,
    fit_anova = fit_anova,
    fit_lme   = fit_lme
  ))
}
```

@@

The function `plot_example()` from the file `plot_example.R` in the folder `scripts/` uses the output of `example.R` to show the fit of a rm-ANOVA and a LMEM. This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case. The corresponding rm-ANOVA and LMEM fits are also presented, and we show the complete output in the next subsection.

@@

## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
plot_example <- function(sim_dat) {
  txt <- 20
  p1 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
    geom_point(show.legend = FALSE) +
    labs(y = 'response')+
    geom_line(aes(x = time, y = mu, color = treatment), 
              show.legend = FALSE) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text=element_text(size = txt)) +
    thm1

  #plot the simulated data with trajectories per each subject
  p2 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
    geom_line(aes(size = "Subjects"), show.legend = FALSE) +
    # facet_wrap(~ treatment) +
    geom_line(aes(x = time, y = mu, color = treatment, 
                  size = "Simulated Truth"),
              lty = 1, show.legend = FALSE) +
    labs(y = 'response') +
    scale_size_manual(name = "Type", 
                      values = c("Subjects" = 0.5, "Simulated Truth" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size = txt)) +
    thm1

  #plot the errors
  p3 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
    geom_line(show.legend = FALSE) +
    labs(y = 'errors') +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size = txt))+
    thm1

  #plot the model predictions for rm-ANOVA
  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
    geom_point(show.legend = FALSE) +
    labs(y = 'response')+
    geom_line(aes(y = predict(sim_dat$fit_anova), 
                  group = subject, 
                  size = "Subjects"),
              show.legend = FALSE) +
    geom_line(data = sim_dat$pred_dat,
              aes(y = predict(sim_dat$fit_anova,
                              level = 0,
                              newdata = sim_dat$pred_dat),
                  size = "Population"),
              show.legend=FALSE) +
    guides(color = guide_legend(override.aes = list(size = 2))) +
    scale_size_manual(name = "Predictions",
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size = txt)) +
    thm1

  #plot the LMEM predictions
  p5 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
    geom_point()+
    labs(y = 'response')+
    geom_line(aes(y = predict(sim_dat$fit_lme),
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat,
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0,
                              newdata = sim_dat$pred_dat),
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions",
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size=txt))+
    thm1

  if(option=='simple'){
    return((p1 + p4 + p5) + plot_layout(nrow = 1) + 
             plot_annotation(tag_levels = 'A'))
  }
  else {
    return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + 
             plot_annotation(tag_levels = 'A'))
  }
}

@@

In this subsection, we use the `example()` and `plot_example()` functions to create an expanded version of Figure 1 in the main manuscript. The difference here is that in the main manuscript `plot_example()` uses `option='simple'` to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit to the simulated data with correlated errors. Here, we use the option `composite` to generate the additional panels for the uncorrelated errors and their respective rm-ANOVA and LMEM fits.

@@

The chunk below sources both `example.R` and `plot_example_Appendix.R` files to simulate data and create the composite plots.

@@

source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
                  "scripts", "example.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
                  "scripts", "plot_example.R"))

A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"),
                 option = 'composite')

B1 <- plot_example(example(fun_type = "linear", error_type = "independent"),
                 option = 'composite')

C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
                 option = 'composite')

D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), 
                 option = 'composite')

@@

```{r, linear-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
# linear response, correlated errors (subject effect)
A1 / B1 + plot_annotation(tag_levels = 'A')
```

@@

```{r, quadratic-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
# linear response, correlated errors (subject effect)
C1 / D1 + plot_annotation(tag_levels = 'A')
```

@@

From the model, we extract the original and weighted four basis functions (and the intercept), and add these last weighted basis functions to construct the smoother.

@@

#basis functions: this script creates Fig 2. by calculating a GAM for the Group 1 data of Figure 1,
# extracting the basis functions
#and creates objects p11,p12,p13,p14 for plotting, which are combined in b_plot to create the
#final composite figure
n_time = 6
x <- seq(1,6, length.out = n_time)
mu <- matrix(0, length(x), 2)
mu[, 1] <- -(0.25 * x^2) + 1.5 * x - 1.25 #mean response
mu[, 2] <- (0.25 * x^2) - 1.5 * x + 1.25 #mean response
y <- array(0, dim = c(length(x), 2, 10))
errors <- array(0, dim = c(length(x), 2, 10))
for (i in 1:2) {     # number of treatments
  for (j in 1:10) {  # number of subjects
    # compound symmetry errors
    errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
    y[, i, j] <- mu[, i] + errors[, i, j]
  }
}

@@

#extract  "Group 1" to fit the GAM
dat <- subset(dat, treatment==1)
#keep just the response and timepoint columns
dat <- dat[,c('y', 'time')]

#GAM model of time, 5 basis functions
gm <- gam(y ~ s(time, k = 5), data = dat, method = "REML")

#model_matrix (also known as) 'design matrix'
#will contain the smooths used to create  model 'gm'
model_matrix <- as.data.frame(predict(gm, type = 'lpmatrix'))

time <- c(1:6)

basis <- model_matrix[1:6, ] #extracting basis (because the values are repeated after every 6 rows)
#basis<-model_matrix[1:6,-1] #extracting basis
colnames(basis)[colnames(basis) == "(Intercept)"] <- "s(time).0"
basis <- basis %>% #pivoting to long format
  pivot_longer(cols = starts_with("s")) %>%
  arrange(name) #ordering

#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
ln <- 6 * (length(coef(gm)))

basis_plot <- data.frame(Basis = integer(ln),
                         value_orig = double(ln),
                         time = integer(ln),
                         coef = double(ln))

basis_plot$time <- rep(time) #pasting timepoints
basis_plot$Basis <- factor(rep(c(1:5), each = 6)) #pasting basis number values
basis_plot$value_orig <- basis$value #pasting basis values
basis_plot$coef <- rep(coef(gm)[1:5], each = 6) #pasting coefficients
basis_plot <- basis_plot %>%
  mutate(mod_val = value_orig * coef) #the create the predicted values the bases need to be
#multiplied by the coefficients

@@

basis_names <- c(`1`="Intercept", `2`="1", `3`="2", `4`="3", `5`="4")

@@

smooth <- basis_plot %>%
  group_by(time) %>%
  summarize(smooth = sum(mod_val))

#original basis
sz <- 1
p11 <- ggplot(basis_plot,
              aes(x = time, y = value_orig, colour = as.factor(Basis))) +
  geom_line(size=sz, show.legend = FALSE) +
  geom_point(size = sz + 1, show.legend = FALSE) +
  labs(y = 'Basis functions') +
  facet_wrap(~ Basis, labeller = as_labeller(basis_names)) +
  theme_classic() +
  thm1

#penalized basis
p12 <- ggplot(basis_plot,
              aes(x = time, y = mod_val, colour = as.factor(Basis))) +
  geom_line(show.legend = FALSE, size = sz) +
  geom_point(show.legend = FALSE, size = sz + 1) +
  labs(y = 'Penalized \n basis functions') +
  scale_y_continuous(breaks = seq(-1, 1, 1)) +
  facet_wrap(~ Basis, labeller = as_labeller(basis_names)) +
  theme_classic() +
  thm1

#heatmap of the  coefficients
x_labels <- c("Intercept", "1", "2", "3", "4")
p13 <- ggplot(basis_plot,
              aes(x = Basis, y = Basis)) +
  geom_tile(aes(fill = coef), colour = "black") +
  scale_fill_gradient(low = "white", high = "#B50A2AFF") +
  labs(x = 'Basis', y = 'Basis') +
  scale_x_discrete(labels = x_labels) +
  geom_text(aes(label = round(coef, 2)),
            size = 7, show.legend = FALSE) +
  theme_classic()+
  theme(legend.title = element_blank())

#plotting simulated datapoints and smooth term
p14<-ggplot(data = dat,
            aes(x = time, y = y)) +
  geom_point(size = sz + 1, alpha = 0.5) +
  thm1 +
  labs(y = 'Simulated \n response') +
  geom_line(data = smooth, aes(x = time, y = smooth),
            color = "#6C581DFF", size = sz + 1)+
  theme_classic()

#Combining all
b_plot <- p11 + p13 + p12 + p14 + plot_annotation(tag_levels = 'A') &
  theme(text = element_text(size=18))

@@

The next chunk presents the code of the function `simulate_data()` from the file `simulate_data.R` in the `scripts/` folder, which we use in the main manuscript to simulate data that follows the reported trends in $\mbox{StO}_2$ in the literature. Although the default standard deviation (SD) to generate the normally-distributed data in the function is 10%, we use a standard deviation of 10% in the main manuscript. We use the default number of 10 observations at each time point in the simulated data. Note that the input for the function is `dat`, needs to be a dataframe that contains the original values for the trends in each group over time from the literature. In our manuscript, we use the dataframe

@@

#data from Vishwanath et. al.
dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56),
              Day = rep(c(0, 2, 5, 7, 10), times = 2),
              Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
```

@@

simulate_data <- function(dat, n = 10, sd = 5) {
  dat_sim <- dat %>%
    slice(rep(1:n(), each = n)) %>%
    group_by(Group, Day) %>%
    mutate(
      StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
      subject = rep(1:10),
      subject = factor(paste(subject, Group, sep = "-"))
    ) %>%
    ungroup()

@@

We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function `confint` methods from the _gratia_ package. The boundaries for the pairwise and simultaneous CIs are obtained by using the type `"confindence"` or `"simultaneous"` in `confint` and are stored in `ci` and `si` respectively.

@@

```{r,pointwise-plot, eval=FALSE}
#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5), 
              data  = dat_missing)
#appraise the model
appraise(mod_m1)

@@

f6 <- ggplot(ci, aes(x = Day, y = est, group = smooth)) +
  geom_line(lwd = 1) +
  geom_ribbon(data = ci, 
              aes(ymin = lower, ymax = upper, x = Day, 
                  group = smooth, fill = Group),
              inherit.aes = FALSE, alpha = 0.7, show.legend = FALSE) +
  geom_ribbon(data = si,
              aes(ymin = lower, ymax = upper, x = Day, 
                  group = smooth, fill =Group),
              inherit.aes = FALSE, alpha = 0.4, show.legend = TRUE) +
  geom_line(data = si, aes(Day, upper, color = Group), 
            size = 0.8, alpha = 0.7) +
  geom_point(data = dat_missing, aes(x = Day, y = StO2_sim, color=Group), 
             size = 1.5, alpha = 0.6, inherit.aes = FALSE) +
  labs(y = expression(atop(StO[2], '(incomplete observations)'))) +
  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
  theme_classic() +
  theme(axis.text=element_text(size = 22)) +
  thm1

@@

This subsection presents the code used in functions `pointwise_comparisons()` and `difference_smooths()` from the files `pointwise_comparisons.R` and `difference_smooths.R` in the `scripts/` folder, which generate the across the function (pointwise) and simultaenous CI for the estimated difference between the smooths. In essence, `pointwise_comparisons()` computes the difference between the smooths and the pointwise CI. Then, `difference_comparisons()` uses the output of `pointwise_comparisons.R` to estimate the simultaneous CI.

@@

Then, `difference_smooths()` function from the file `difference_smooths.R` in the `scripts/` folder takes the output from the previous function and uses it to compute the simultaneous CI.
We present the code below.

@@

  # smooths in model
  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
  # select smooths
  select <-
    gratia:::check_user_select_smooths(smooths = S, select = smooth,
                                       partial_match = partial_match)

  sm_ids <- which(select)
  smooths <- gratia::get_smooths_by_id(model, sm_ids)
  sm_data <- map(sm_ids, gratia:::smooth_data,
                 model = model, n = n, include_all = TRUE)
  sm_data <- bind_rows(sm_data)
  by_var <- by_variable(smooths[[1L]])
  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
                                   stringsAsFactor = FALSE))
  names(pairs) <- paste0("f", 1:2)

@@

pairwise_limits <- function(dataframe) {
  #extract values where the lower limit of the ribbon is greater than zero
  #this is the region where the control group effect is greater

  v1 <- dataframe %>%
    filter(lower_s > 0) %>%
    select(Day)
  #get day  initial value
  init1 = v1$Day[[1]]
  #get day final value
  final1 = v1$Day[[nrow(v1)]]
  #extract values where the value of the upper limit of the ribbon is lower than zero
  #this corresponds to the region where the treatment group effect is greater
  v2 <- dataframe %>%
    filter(upper_s < 0) %>%
    select(Day)

  init2 = v2$Day[[1]]
  final2 = v2$Day[[nrow(v2)]]
  #store values
  my_list <- list(init1 = init1,
                  final1 = final1,
                  init2 = init2,
                  final2 = final2)
  return(my_list)
